Accelerated convergence of convective simulations using boundary value problems

### 2017-12-15 tex/bvp_paper.tex
adds plot script for parameter space comparison plot to repo

\section{Introduction}
\label{sec:intro}
In numerical studies of convection, the chosen boundary conditions at the upper and lower plates
determines key quantities of the dynamics of the evolved state.
Studies of incompressible, Boussinesq, \RB convection (RBC) often
employ fixed temperature (Dirichlet) or fixed heat flux
(Neumann) boundary conditions at both plates.  
Dirichlet conditions represent plates of infinite conductivity,
whereas Neumann condition model plates of finite conductivity.  
In both cases, choosing symmetric boundary conditions maintains overall system symmetry, 
and despite evolving towards quite different thermal structures, both types of conditions
transport heat in the same manner \cite{johnston&doering2009}.

Studies of convection in stratified systems which aim to model convection 
in astrophysical systems, such as the outer envelopes of low-to-moderate mass stars 
like the Sun, often employ a mixture of these
two types of boundary conditions \cite{hurlburt&all1984, cattaneo&all1991}.  
The flux at the lower boundary is fixed, modeling
the constant energy generation of the stellar core, 
while the outer boundary condition is held at a fixed temperature,
modeling the surface of a star which must output the energy generated internally.
This set of boundary conditions is also being studied more recently in Boussinesq convection
\cite{korre&all2017}, and is displaying interesting characteristics such as internal layers which
are stable to convection.

While this setup is a useful model for understanding natural
systems, simulations which employ this setup often suffer from a long convective transient as the thermodynamic structure
of the atmosphere relaxes to the adiabatic profile specified by the fixed temperature upper boundary condition.
This long evolution occurs on the ``Kelvin-Helmholtz,'' or thermal diffusion timescale of the atmosphere, 
$t_{\text{KH}} \approx L_z^2 / \chi$, where $L_z$ is the domain depth and $\chi$ is the thermal diffusivity.
Interesting convection happens as high values of the Rayleigh number, which scales like $\chi^{-1/2}$, such that
in the astrophysically interesting regime of high-Ra, highly stratified convection, evolving a simulation for a
KH time becomes numerically intractable.  As the Rayleigh number increases, the KH time increases while the average
timestep required to resolve the more turbulent flows decreases.  The net result is that under standard initial conditions
of hydrostatic- and thermal- equilibrium, the desired convective solution cannot be obtained and the dynamics of convection
there cannot be studied.

Knowledge of Mixing Length theory and the nature of evolved convection has been used in some 
previous studies (e.g., \cite{brandenburg&all2005}) to choose initial conditions which are
closer to the correct evolved adiabat than
the hydrostatic polytropic states.  However, these assumptions work best in convective regions
which are bounded by stable layers.  In simple atmospheres where boundary layers form to meet
boundary conditions at the upper and lower edge of the atmosphere, we cannot know 
\emph{a priori} what the extent or shape of the boundary layer will be, as that must be chosen
by the convective dynamics, and specification of the proper boundary layer is essential for
placing the atmosphere along the proper adiabat.

Here we present a method for using simple boundary value problems, 
along with information about the evolved flow fields,
to fast-forward the slow thermal evolution of convecting simulations.  
We present this method in the context of RBC, and
then demonstrate that applying it to stratified, compressible convection 
is simple.  These methods allow us to study the convective flows driven by 
nearly the evolved thermal profile while only requiring initial value problems to run
for long enough to resolve the fast dynamical timescales of convection.

\section{Incompressible, Boussinesq \RB Convection}
\label{sec:experiment}
In general, the method for fast-forwarding the atmospheric structure to the correct adiabat is
simple and involves only a few steps.  At the start of convective simulations, a large transient
occurs while the hydrostatic state gives way to the convective state.  After the peak of this
transient, we begin to average the mean vertical profile of the convective flux and the conductive
flux through the domain.  Once those flux profiles are converged, we solve a simple 
boundary value problem using the information that those fluxes provide to determine what the
evolved thermal structure of the atmosphere can be.  The stratification of the initial value
problems is then adjusted to match the output of the BVP, and we run convection for dynamical
timescales to let the flows adjust to the new, updated thermal structure.  We then measure
quantities of the dynamics, and compare them to the dynamics of simulations which underwent a
long thermal rundown, to determine if our simple BVPs effectively get our convective solution
to the correct state.

\subsection{Governing Equations, Nondimensionalization, and Domain setup}
In our first set of experiments, we adopt the Oberbeck-Boussinesq approximation.  Here, the
fluid has constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$).  We non-dimensionalize length by the layer height ($L_z$) ,
temperature by the (constant) initial temperature gradient across the layer ($\grad T_0$), and time
by the freefall timescale ($L_z / v_{\text{ff}}$, with $v_{\text{ff}} = \sqrt{\alpha g L_z^2 \grad T_0}$, where $g$ is 
uniform gravitational acceleration in the $-\hat{z}$ direction). The dimensionless Boussinesq
equations governing the velocity $\bm{u} = u\hat{x} + v\hat{y} + w\hat{z}$, temperature
$T = T_0 + T_1$, and reduced pressure $\varpi$ are \cite{spiegel&veronis1960}
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy}
\end{gather}
where the dimensionless control parameters are the Rayleigh and Prandtl numbers,
\begin{equation}
\text{Ra} = \frac{g \alpha L^4 \left(\frac{dT}{dz}\right)_0}{\nu\chi} = \frac{(L\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}.
\end{equation}
The dimensionless vertical extent of the domain is $z = [-1/2, 1/2]$, and at the boundaries
we impose no-slip, impenetrable boundary conditions such that $w = u = v = 0$ at $z = \pm 1/2$.
At the lower boundary, we employ a fixed flux condition such that $\partial T_1 / \partial z = 0$
at $z = -1/2$, and we impose a fixed temperature condition $T_1 = 0$ at $z = 1/2$. Both
horizontal directions are periodic and have equal aspect ratio, $\gamma = 4$, such that
the horizontal coordinates $x, y = [0, \Gamma]$.

\subsection{The Boundary Value Problem}
The Boussinesq BVP is simply the time-stationary, horizontally- and time- averaged equations
of hydrostatic balance and energy conservation,
\begin{gather}
\frac{\partial}{\partial z}\angles{\varpi} = \angles{T_1}\hat{z},
	\label{eqn:bouss_BVP_momentum}
\\
\frac{\partial}{\partial z}\angles{wT_1} = \frac{1}{\text{Pr}\text{Ra}}\frac{\partial^2}{\partial z^2} \angles{T_1},
	\label{eqn:bouss_BVP_energy}
\end{gather}
where $\angles{A}$ represents a time- and horizontally averaged profile of the quantity $A$.  These
equations arise from taking time- and horizontal- averages of Eqns (\ref{eqn:bouss_momentum_eqn}-\ref{eqn:bouss_energy_eqn})
and neglecting terms that vanish due to symmetry in the evolved flows.  Convective flows
are perturbations around a thermal profile defined by these equations in the proper evolved state.

@@

>>>>

\section{Introduction}
\label{sec:intro}
Natural convection occurs in the presence of disparate timescales. Granules on the
solar surface overturn on the order of 10 minutes, whereas deep motions in the Sun are likely at
low Mach number and constrained by the solar rotation rate of \~1 month.  
Despite these relatively short dynamical times, the scale of energy transport on the
Sun occurs on the Kelvin-Helmholtz timescale of nearly $3 \times 10^7$ \emph{years} \cite{stix2003}. 
As simulations aim to model natural convection
by increasing into the high-Rayleigh Number (Ra) regime, where diffusive timescales are much
longer than dynamical timescales \cite{Anders&Brown2017}, achieving converged simulations will 
require runs which span a greater number of convective timescales in order to thermally converge.
Furthermore, with increasing Ra and decreasing diffusivities, motions become more turbulent
and require finer grid meshes and smaller timesteps to resolve turbulent motions, meaning that
achieving even one overturn timescale becomes a harder problem.  
These two effects combine to make thermally converging high-Ra, astrophysically interesting
simulations intractable using modern numerical tools.

In studies of stratified convection where a convective layer lies between stable layers, studies
have used the knowledge of Mixing Length Theory (MLT) to adjust the initial thermal profile of
atmospheres to a state which is closer to the adiabat chosen by convection \cite{brandenburg&all2005}.
However, many studies of convection do not contain stable layers above and below the convection
zone, and the presence of hard boundaries and the boundary layers that they form means
that the proper adiabat cannot be known \emph{a priori} until the simulation evolves the
structure of the boundary layers.

The chosen boundary conditions at the upper and lower plates
determine key quantities of the dynamics of the evolved state.
Studies of incompressible, Boussinesq, \RB convection (RBC) often
employ fixed temperature (Dirichlet) or fixed heat flux
(Neumann) boundary conditions at both plates.  
Dirichlet conditions represent plates of infinite conductivity,
whereas Neumann conditions model plates of finite conductivity.  
In both cases, choosing symmetric boundary conditions maintains overall system symmetry, 
and despite evolving towards different thermal structures, both types of conditions
transport heat in the same manner \cite{johnston&doering2009}.

Studies of convection which aim to model
in astrophysical systems, such as the outer envelopes of low-to-moderate mass stars 
like the Sun, often employ a mixture of these
two types of boundary conditions \cite{hurlburt&all1984, cattaneo&all1991, korre&all2017}.  
The flux at the lower boundary is fixed, modeling
the constant energy generation of the stellar core, 
while the outer boundary condition is held at a fixed temperature,
modeling the surface of a star which must output the energy generated internally.
While this setup is a useful model for understanding natural
systems, simulations which employ this setup often suffer from a long 
thermal relaxation as the atmosphere loses energy and approaches the adiabat chosen by the
Dirichlet condition.

Here we present a method for using simple boundary value problems (BVPs), 
along with information about the evolved flow fields,
to fast-forward the slow thermal evolution of convecting simulations.  
We run two sets of experiments: one in which we allow convective simulations to evolve for a
full thermal timescale before taking measurements, and another in which we employ a fast-forwarding,
BVP technique which occurs on dynamical timescales. We compare these two sets of simulations to
show the validity of the BVP technique.  Then, we use the BVP technique to run simulations
at high Ra, in the regime where running for thermal timescales becomes computationally intractable.

\section{Experiment}
\label{sec:experiment}
In our study, we adopt the Oberbeck-Boussinesq approximation.  Here, the
fluid has constant kinematic viscosity ($\nu$), thermal diffusivity ($\kappa$), and coefficient
of thermal expansion ($\alpha$).  We non-dimensionalize length by the layer height ($L_z$) ,
temperature by the (constant) initial temperature gradient across the layer ($\grad T_0$), and time
by the freefall timescale ($L_z / v_{\text{ff}}$, with $v_{\text{ff}} = \sqrt{\alpha g L_z^2 \grad T_0}$, where $g$ is 
uniform gravitational acceleration in the $-\hat{z}$ direction). The dimensionless Boussinesq
equations governing the velocity $\bm{u} = u\hat{x} + v\hat{y} + w\hat{z}$, temperature
$T = T_0 + T_1$, and reduced pressure $\varpi$ are \cite{spiegel&veronis1960}
\begin{gather}
\DivU = 0, 
	\label{eqn:incompressible}
\\
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} =
-\grad\varpi + T_1\hat{z} + \frac{\text{Pr}}{\text{Ra}}\grad^2\bm{u}, 
	\label{eqn:bouss_momentum}
\\
\frac{\partial T_1}{\partial t} + \bm{u}\cdot\grad(T_0 + T_1) = \frac{1}{\text{Pr}\,\text{Ra}}\grad^2 T_1,
	\label{eqn:bouss_energy}
\end{gather}
where the dimensionless control parameters are the Rayleigh and Prandtl numbers,
\begin{equation}
\text{Ra} = \frac{g \alpha L^4 \left(\frac{dT}{dz}\right)_0}{\nu\chi} = \frac{(L\,v_{\text{ff}})^2}{\nu\chi}, \qquad \text{Pr} = \frac{\nu}{\chi}.
\end{equation}
The dimensionless vertical extent of the domain is $z = [-1/2, 1/2]$, and at the boundaries
we impose no-slip, impenetrable boundary conditions such that $w = u = v = 0$ at $z = \pm 1/2$.
At the lower boundary, we employ a fixed flux condition such that $\partial T_1 / \partial z = 0$
at $z = -1/2$, and we impose a fixed temperature condition $T_1 = 0$ at $z = 1/2$. Both
horizontal directions are periodic and have equal aspect ratio, $\gamma = 4$, such that
the horizontal coordinates $x, y = [0, \Gamma]$.

\subsection{The Boundary Value Problem}
The Boussinesq BVP is simply the time-stationary, horizontally- and time- averaged equations
of hydrostatic balance and energy conservation,
\begin{gather}
\frac{\partial}{\partial z}\angles{\varpi} = \angles{T_1}\hat{z},
	\label{eqn:bouss_BVP_momentum}
\\
\frac{\partial}{\partial z}\angles{wT_1} = \frac{1}{\text{Pr}\text{Ra}}\frac{\partial^2}{\partial z^2} \angles{T_1},
	\label{eqn:bouss_BVP_energy}
\end{gather}
where $\angles{A}$ represents a time- and horizontally averaged profile of the quantity $A$.  These
equations arise from taking time- and horizontal- averages of Eqns (\ref{eqn:bouss_momentum}-\ref{eqn:bouss_energy})
and neglecting terms that vanish due to symmetry in the evolved flows.  Convective flows
are perturbations around a thermal profile defined by these equations in the proper evolved state.

@@

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/parameter_space_comparison.png}
\caption{\label{fig:parameter_space_comparison} }
\end{figure}
